Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology

### 2024-04-26 analysis/supplementary-materials/SOM_lithotypes.qmd
Pre submission commit_1

---
title: "Terrace and shelter lithotype complete description"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
title-block-published: "Last updated"  
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: ../paper/references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  out.width = "100%",
  fig.align = "center"
)
```

# Chert group description

## Group 2

Group 2 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Pliensbachian.

It is a fine-grained chert, with a conchoidal fracture and good quality (see the Data dictionary for macroscopic analysis document for details on the attribution of quality to raw materials used in this analysis). It is macroscopically heterogeneous in color, patterns, and textural characteristics; due to this, initially, two sub-groups were created: single-color @fig-type2-variability and multiple-color () groups.

The single-color group's main colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), grayish brown (10YR 5/2) and white (2.5Y 8/1). The color variability and presence of allochems and bioclasts produce macroscopic patterns such as splotched, speckling, and laminations. The structure is homogeneous. The artifacts have medium to dull luster; some samples show shiny luster, however, this may be related to alterations to the surface. Translucency is opaque and the feel is smooth to semi-smooth. The multiple-color group shows a variable color distribution: colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), pale red (2.5YR 6/2), and white (2.5Y 8/1). The color variability creates macroscopic patterns such as splotched, broad mottling, speckling, and laminations (Fig. \@ref(fig:type2Bvariability)). The structure is similar to group single-color, although slightly heterogeneous as translucency varies from opaque to sub-translucent, with light passing through at around 3.5 mm, limited to the white areas.

For both sub-groups, whenever present, the cortex can be rounded or irregular and from an outcrop, varying between ++ to ++ mm in thickness, frequently with a sharp transition.

Inclusions include black speckles, which may be organics or iron oxides, and red speckles which may be iron oxides. Bioclasts include white, round speckles, spicule-shaped skeletal remains, monaxon sponge spicules, bivalve shells (rare, hereon r) replaced by quartz or chalcedony, and rarely triaxon sponge spicules. In the single-color group fossil remains range from uncommon to frequent, while in the multiple-color group fossils range from uncommon to very frequent; in all cases, the bioclasts are frequently unidentifiable.

In the shelter area, the variety of identified fossils also includes possible ostracods (uncommon, hereon u), a possible small foraminifer (Fig. \@ref(fig:shelter2fossils) d), and echinoderm spines (Fig. \@ref(fig:shelter2fossils) a-c).

Five thin sections were produced to account for the macroscopic variability of this group. All thin sections show a wackestone texture and massive microstructure. Micro-cryptocrystalline quartz represents more than 90% of the orthochem composition (between 93 to 99%), while other orthochems are present in lower frequencies: macrocrystalline quartz (1-5%) as grains scattered in the sample or replacing fossils and fibrous chalcedony (1-5%) replacing fossils. Accessory minerals are also present in some samples: dolomite, micrite, and mica (muscovite). Allochems present are oxide patina and grains (common, hereon c), unidentifiable opaques (common to very frequent, hereon vf), and bioclasts (vf). All samples have very frequent unidentifiable fossils. In samples from multiple-color group, although in lesser frequencies, monaxon sponge spicules (H21-3095) and a possible triaxone sponge spicule (H21-4234) are identifiable. All samples show vug-type porosity between 1-5%.

```{r}
#| label: fig-type2-variability
#| fig-cap: Chert macroscopic variability within group 2 (single color).
#| fig-width: 8

knitr::include_graphics("../figures/SOM/Group2_macromeso_b.png")

@@

```{r type2Bvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 2 (multiple color).", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group2B_macromeso.png")

@@

```{r shelter2fossils, echo=FALSE, fig.cap="Identified fossil variability from group 2 from the Shelter area. Fossils present in the figures are: a-b, crinoid stem section with 4 series; c, cross-section of crinoid columnal with pentagonal stem plate; d, possible foraminifer (black arrow).", out.width = '80%'}

@@

A group of samples was individualized from group 2 due to color and textural differences (named 2R during the analysis process). These are interpreted as alterations to group 2 cherts, possibly due to heat, although further experimentation is needed. The color distribution is variable and it has a heterogeneous structure, possibly due to the alterations. The colors are red (10R 5/6), light yellowish brown (10YR 6/4), and pinkish-gray (5YR 7/2). It shows shiny to dull luster and semi-smooth to rough feel.

@@

```{r type2Rvariabilitymeso, echo=FALSE, fig.cap="Archaeological samples attributed to Type 2R. ", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group2R_macromeso-01.png")

@@

A thin section was made to confirm whether the color and textural differences are solely macroscopic, and do not represent a different petrographic type. The sample (H19-2426 from Level 4D) shows no outstanding differences from the samples from group 2.

## Group 15

Group 15 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Pliensbachian.

Group 15 is a fine to medium chert, with a conchoidal fracture and good quality. The color distribution is uniform, despite subtle changes to gray. The colors are light gray (10R 7/1) and reddish gray (10YR 6/1). It has medium to dull luster and is sub-translucent to opaque, with light passing through only in the very fine parts (\~1.4 mm). The splotches and lamination adjacent to the cortex seem to be sub-translucent as well, showing more translucency than the chert itself. The feel is semi-smooth and the patterns are splotches, speckling, concentric lamination (limited to a small number of samples), and fine lamination (found only between the cortex and the chert).

Whenever present, the cortex seems to be from an outcrop and rounded. Despite the brown coloration possibly due to post-depositional alterations, the original color is white. Cortex thickness ranges from 3.8 mm to 1.2 mm, and has a sharp transition. One single sample from Level 6 shows a gradual cortex transition. Although the original morphology of this chert is unknown, the presence of cortex on two opposed extremities in a sample (H19-3196, \@ref(fig:type15variability) a) indicates the chert can be found in irregularly rounded nodules within the parent rock.

@@

```{r type15variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 15.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group15_macromeso.png")
```

## Group 7

Group 7 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin locate their epoch and stage in the Lower Jurassic, Pliensbachian.

@@

Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: grey (G) and white (W) fabric. The grey fabric corresponds to the interior of the nodule. It is light gray (5Y 7/1, 7/2), light olive gray (5Y 6/2), and gray (5Y 6/1, 5/1), sub-translucent (light passing through in the thin areas, around 3.2 mm thickness) and has medium luster. The visible inclusions are limited to a single mica mineral (\~50 μm), white flecks (concentrated near the outer edges of white fabric), and speckles, which may be circular, unidentifiable fossil remains, and rare spicule-like remains. In layer 6 spicule-shaped remains are uncommon. The white fabric appears in two areas in the nodules: 1) in the outer edge, surrounding the grey fabric; 2) within the grey fabric, where it creates splotched and broad mottling patterns. This fabric is white (5Y 8/1, 10YR 8/1) and very pale brown (10YR 8/2, 8/4), opaque and dull. In the pale brown splotches, inclusions are yellow speckles (possibly iron oxides, Fig. \@ref(fig:type7variability) a). The white fabric generally has barely visible fossils and uncommon unidentifiable fossils replaced by chalcedony or quartz.

Whenever present, the cortex seems to be from an outcrop, mostly rounded, although a sample from layer 5 shows a thick (30 mm) irregular cortex. The cortex is white or light brown with a sharp transition between 1.4 mm to 10 mm thick. The morphology of the cortex in fully cortical samples (i.e., I18-2099; \@ref(fig:type7variability) d) may indicate that this chert is originally available as irregular (slightly oval) nodules within the parent rock.

Petrography results show this group is a massive wackestone composed mostly of micro-cryptocrystalline quartz (98%), with accessory macrocrystalline quartz and fibrous chalcedony. Allochems are iron oxides (c), opaques (vf), and bioclasts (c). Despite the macroscopically visible spicule-like white shapes, these fossils are not apparent in thin section. The porosity is low (1%).

```{r type7variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 7.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group7_macromeso.png")

@@

## Group 12

Group 12 corresponds to a massive and banded micro-cryptocrystalline quartz from a marine environment and possibly a shelf depositional environment. The macroscopic and petrographic association of this group with the local cherts from the west Algarve basin, as well as the large size of the nodules and presence of parent rock, allow us to interpret this chert as local, possibly the Lower Jurassic although the exact outcrop is not identifiable.

This chert is fine to medium-grained, with a conchoidal fracture (albeit uneven at times). The quality is acceptable to low due to the lack of homogeneity and frequent fractures, as well as a coarser fabric in certain areas. It has a heterogeneous structure, frequently showing different characteristics within one single artifact. The colors are gray (5Y 6/1), light brownish gray (10YR 6/2), and pale brown (10YR 6/3). Fresh surfaces display slightly different colors: light gray (2.5Y 7/1) and gray (2.5Y 6/1). It has medium to dull luster, coarse to medium feel, and is opaque to sub-translucent; in the sub-translucent areas, light passes through to max. 4 mm of thickness. The group displays frequent patterns, mostly related to this heterogeneity, but also due to patination; these are splotched, banded, broad mottling, and speckling.

The cortex is from an outcrop, frequently rounded, but also irregular. Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules (Fig. \@ref(fig:type12variability) b). The cortex is thin (\~0.5 mm), with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex (Fig. \@ref(fig:type12variability) a).

Based on the presence of bands, two sub-groups were created: 12 (Fig. \@ref(fig:type12variability) a-c) and 12B (Fig. \@ref(fig:type12variability) d), the latter characterized by banding. Two thin sections were produced to characterize the macroscopic variability. The petrographic results show that group 12 is a wackestone, with a massive (group 12) or banded (group 12B) microstructure. They are mainly composed of micro-cryptocrystalline quartz (75% and 72%, respectively) and micrite (20% in both), and in smaller percentages macrocrystalline quartz and fibrous chalcedony. Allochems are unidentified opaques (vf), iron oxides (u), and bioclasts (c). The sub-groups show larger differences in the porosity, with 12 showing 1% vug porosity while 12B shows 10% vug and moldic porosity, concentrated in the banded areas with more opaques.

Bioclasts are uncommon both macroscopically and in thin section: sponge spicules and round, unidentifiable fossils can be seen by the naked eye or under the microscope. In thin section however, several other unidentified fossils are present, some of which may be an ostracod (both in 12 and 12B) and a foraminifer (12).

```{r type12variability, echo=FALSE, fig.cap="Chert macroscopic variability within group 12.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/Group12_macromeso.png")

@@

## Group 11

Group 11 corresponds to a massive cryptocrystalline quartz with mudstone texture. The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage. However, rare spicules were identified; given the exclusivity of sponge's (Porifera) living environment to be marine [@flugelMicrofaciesCarbonateRocks2010], group 11 may be considered a chert formed in a marine environment.

This chert is fine-grained and translucent (light passing through at 17 mm of thickness), with a conchoidal fracture and good quality. The structure is heterogeneous, with shiny to medium luster and semi-smooth to smooth feel. This group of translucent cherts shows a variety of colors: grayish brown (2.5Y 5/2), light brownish gray (10YR 6/2), brown (7.5 YR 4/2), reddish brown (2.5YR 4/3) and dark gray (10YR 4/1). The color distribution is variable; in some cases, the color seems to be related to the thickness of the sample; thicker parts show darker colors, while thinner parts show lighter colors. Alterations and inclusions also impact the color distribution and the discernment of natural versus the altered color. Alterations such as patina also seem to impact the translucency, which becomes opaque, and the luster, which becomes medium to dull.

@@

The cortex is rounded, from an outcrop source, and rarely rounded and smooth like a pebble. It is frequently around 0.5 to ++ mm thick and the transition is gradual.

```{r typeCPTvariability, echo=FALSE, fig.cap="Chert macroscopic variability within group 11.", out.width = '80%'}

knitr::include_graphics("../figures/SOM/GroupCPT_macromeso.png")

@@

### Sub-group LLB

A sub-group named LLB (corresponding to the light brownish gray colors) was created during the macroscopic analysis process due to the presence of a group of artifacts with homogeneous characteristics (and refits) within the previous group.

This sub-group is characterized by white (10YR 8/1), light brownish gray (10YR 6/2), and brown (7.5YR 5/2) colors. It has a heterogeneous structure, with two fabrics: white and brown. The white fabric is located by the cortex and is opaque. It gradually turns to a light brownish gray which is sub-translucent (light passing through until \~10 mm), and to a brown which is translucent. The luster ranges from medium to dull, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is good. Patterns are laminated, splotched, and speckled. The patterns occur on the brown fabric, in a light brownish-gray color.

@@

```{r typeLLB11Evariability, echo=FALSE, fig.cap="Chert macroscopic variability within the sub-group LLB (a-b) and 11E (c-d).", out.width = '100%'}

knitr::include_graphics("../figures/SOM/groupLLB_11E_variability.png")

@@

### Sub-group 11E

Similarly to the previous sub-group, a sub-group named 11E was created during the macroscopic analysis process due to the presence of a group of artifacts with homogeneous characteristics (and refits).

This sub-group is characterized by a heterogeneous structure. The colors are brown (10YR 5/3), gray (2.5Y 5/1), light gray (10YR 7/2) and white (10YR 8/1) The translucency is variable, some areas are translucent (light passing through at 8 mm of thickness), while the splotches are sub-translucent. The luster ranges from shiny to medium, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is acceptable, due to the presence of fractures. Patterns are splotches, broad mottling, spots, speckling, and laminated.

@@

Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups.The results show no outstanding differences between the samples and sub-groups. The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure. It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples, with one exception where it represents 91%), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz and micrite/sparite. Allochems are opaques (u), iron oxides (u) and bioclasts (r/u). Oxides and opaques are often concentrated in the edges of the sample, and may be related to post-depositional processes. Bioclasts are unidentifiable and poorly preserved, replaced by fibrous chalcedony and rarely by microcrystalline quartz. In a single sample (I20-3689) sponge spicules were identified. However, in sample H19-4074 (sub-type 11E), the absence of spicules or spicule-like shapes in the thin section suggests that they are an effect of white patina on the surface of the samples. Porosity is present between \<1-1%, of vug-type when identifiable.

## Group 11F

Group 11F corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment. Identified and well-preserved fossils are foraminifera from the Pfenderenidae family, which are larger benthic foraminifera from marine environments. The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous.

>>>>

---
title: "Chert use and procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia, Journal of Paleolithic Archaeology"
subtitle: "Online Resource 2: Terrace and shelter lithotypes complete description"
author:
  - Joana Belmiro:
      orcid: https://orcid.org/0000-0003-4951-2907
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute: 
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - Salvador Dominguez-Bella:
      orcid: https://orcid.org/0000-0003-3892-763X
      institute: 
        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated"  
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: ../paper/references.bib
toc: true
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  dpi = 300,
  out.width = "70%",
  fig.align = "center"
)
```

# Chert type description

## Type 1

Type 1 corresponds to a hydrothermal, fibrous chalcedony. The macroscopic and petrographic association of this type with the local Triassic chalcedonies from the west Algarve basin locate its epoch in the Triassic.

It is a fine-grained chert, with a conchoidal fracture and acceptable quality, due to the frequent fractures and porosity within the lenses and nodules (see the Data dictionary for macroscopic analysis document for details on the attribution of quality to raw materials used in this analysis). It is homogeneous, without color but frequently white patinated. The artifacts are dull (or with a waxy appearance), translucent (whenever not patinated) and smooth to semi-smooth.

Cortex is difficult to identify due to the frequent patina, fractures and porosity, although whenever present, it is thin. Although found in lenses in the outcrops, the artifacts do not display any tabular characteristics (cortex in two opposed faces).

Results from thin-section show this chert is composed by 100% fibrous chalcedony, without allochems or bioclasts.

```{r}
#| label: fig-type1-variability
#| fig-cap: Chert macroscopic variability within type 1 (chalcedony).

knitr::include_graphics("../figures/SOM/type1_macromeso.png")

@@

## Type 2

Type 2 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment (@fig-type2-variability and @fig-type2Bvariability). The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Carixian.

It is a fine-grained chert, with a conchoidal fracture and good quality. It is macroscopically heterogeneous in color, patterns, and textural characteristics; due to this, initially, two sub-types were created: single-color (@fig-type2-variability) and multiple-color (@fig-type2Bvariability) types.

The single-color type's main colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), grayish brown (10YR 5/2) and white (2.5Y 8/1). The color variability and presence of allochems and bioclasts produce macroscopic patterns such as splotched, speckling, and laminations. The structure is homogeneous. The artifacts have medium to dull luster; some samples show shiny luster, however, this may be related to alterations to the surface. Translucency is opaque and the feel is smooth to semi-smooth. The multiple-color type shows a variable color distribution: colors are very pale yellow (10YR 8/4), yellow (10YR 8/6), pale red (2.5YR 6/2), and white (2.5Y 8/1). The color variability creates macroscopic patterns such as splotched, broad mottling, speckling, and laminations (@fig-type2Bvariability). The structure is similar to type single-color, although slightly heterogeneous as translucency varies from opaque to sub-translucent, with light passing through at around 3.5 mm, limited to the white areas.

For both sub-types, whenever present, the cortex can be rounded or irregular and from an outcrop, varying between 10 to 0.5 mm in thickness, frequently with a sharp transition.

Inclusions include black speckles, which may be organics or iron oxides, and red speckles which may be iron oxides. Bioclasts include white, round speckles, spicule-shaped skeletal remains, monaxon sponge spicules, bivalve shells (rare, hereon r) replaced by quartz or chalcedony, and rarely triaxon sponge spicules. In the single-color type fossil remains range from uncommon to frequent, while in the multiple-color type fossils range from uncommon to very frequent; in all cases, the bioclasts are frequently unidentifiable.

In the shelter area, the variety of identified fossils also includes possible ostracods (uncommon, hereon u), a possible small foraminifer (@fig-shelter2fossils d), and echinoderm spines (@fig-shelter2fossils a-c).

Five thin sections were produced to account for the macroscopic variability of this type. All thin sections show a wackestone texture and massive microstructure. Micro-cryptocrystalline quartz represents more than 90% of the orthochem composition (between 93 to 99%), while other orthochems are present in lower frequencies: macrocrystalline quartz (1-5%) as grains scattered in the sample or replacing fossils and fibrous chalcedony (1-5%) replacing fossils. Accessory minerals are also present in some samples: dolomite, micrite, and mica (muscovite). Allochems present are oxide patina and grains (common, hereon c), unidentifiable opaques (common to very frequent, hereon vf), and bioclasts (vf). All samples have very frequent unidentifiable fossils. In samples from multiple-color type, although in lesser frequencies, monaxon sponge spicules (H21-3095) and a possible triaxone sponge spicule (H21-4234) are identifiable. All samples show vug-type porosity between 1-5%.

```{r}
#| label: fig-type2-variability
#| fig-cap: Chert macroscopic variability within type 2 (single color).

knitr::include_graphics("../figures/SOM/type2_macromeso.png")

@@

```{r}
#| label: fig-type2Bvariability
#| fig-cap: Chert macroscopic variability within type 2 (multiple color).

knitr::include_graphics("../figures/SOM/type2B_macromeso.png")

```{r}
#| label: fig-shelter2fossils
#| fig-cap: "Identified fossil variability from type 2 from the Shelter area. Fossils present in the figures are: a-b, crinoid stem section with 4 series; c, cross-section of crinoid columnal with pentagonal stem plate; d, possible foraminifer (black arrow)."

@@

A type of samples was individualized from type 2 due to color and textural differences (named 2R during the analysis process; @fig-type2Rvariabilitymeso). These are interpreted as alterations to type 2 cherts, possibly due to heat, although further experimentation is needed. The color distribution is variable and it has a heterogeneous structure, possibly due to the alterations. The colors are red (10R 5/6), light yellowish brown (10YR 6/4), and pinkish-gray (5YR 7/2). It shows shiny to dull luster and semi-smooth to rough feel.

@@

```{r}
#| label: fig-type2Rvariabilitymeso
#| fig-cap: Archaeological samples attributed to Type 2 with reddening alterations.

knitr::include_graphics("../figures/SOM/type2R_macromeso-01.png")

@@

A thin section was made to confirm whether the color and textural differences are solely macroscopic, and do not represent a different petrographic type. The sample (H19-2426 from Level 4D) shows no outstanding differences from the samples from type 2.

## Type 3

Type 3 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment (@fig-type3variability). The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin as well as the identified fossils locate their epoch and stage in the Lower Jurassic, Carixian.

Type 3 is a fine to medium chert, with a conchoidal fracture and good quality. The color distribution is uniform, despite subtle changes to gray. The colors are light gray (10R 7/1) and reddish gray (10YR 6/1). It has medium to dull luster and is sub-translucent to opaque, with light passing through only in the very fine parts (\~1.4 mm). The splotches and lamination adjacent to the cortex seem to be sub-translucent as well, showing more translucency than the chert itself. The feel is semi-smooth and the patterns are splotches (@fig-type3variability d), speckling, concentric lamination (limited to a small number of samples), and fine lamination (found only between the cortex and the chert).

Whenever present, the cortex seems to be from an outcrop and rounded. Despite the brown coloration possibly due to post-depositional alterations, the original color is white (@fig-type3variability a-c). Cortex thickness ranges from 3.8 mm to 1.2 mm, and has a sharp transition. One single sample from Level 6 shows a gradual cortex transition. Although the original morphology of this chert is unknown, the presence of cortex on two opposed extremities in a sample (H19-3196, @fig-type3variability a) indicates the chert can be found in irregularly rounded nodules within the parent rock.

@@

```{r}
#| label: fig-type3variability
#| fig-cap: Chert macroscopic variability within type 3.

knitr::include_graphics("../figures/SOM/type3_macromeso.png")
```

## Type 4

Type 4 corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf depositional environment (@fig-type4variability). The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin locate their epoch and stage in the Lower Jurassic, Carixian.

@@

Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: grey and white fabric. The grey fabric corresponds to the interior of the nodule. It is light gray (5Y 7/1, 7/2), light olive gray (5Y 6/2), and gray (5Y 6/1, 5/1), sub-translucent (light passing through in the thin areas, around 3.2 mm thickness) and has medium luster. The visible inclusions are limited to a single mica mineral (\~50 μm), white flecks (concentrated near the outer edges of white fabric), and speckles, which may be circular (@fig-type4variability b-d), unidentifiable fossil remains, and rare spicule-like remains. In layer 6 spicule-shaped remains are uncommon. The white fabric appears in two areas in the nodules: 1) in the outer edge, surrounding the grey fabric; 2) within the grey fabric, where it creates splotched and broad mottling patterns. This fabric is white (5Y 8/1, 10YR 8/1) and very pale brown (10YR 8/2, 8/4), opaque and dull. In the pale brown splotches, inclusions are yellow speckles (possibly iron oxides; @fig-type4variability a). The white fabric generally has barely visible fossils and uncommon unidentifiable fossils replaced by chalcedony or quartz.

Whenever present, the cortex seems to be from an outcrop, mostly rounded, although a sample from layer 5 shows a thick (30 mm) irregular cortex. The cortex is white or light brown with a sharp transition between 1.4 mm to 10 mm thick. The morphology of the cortex in fully cortical samples (i.e., I18-2099; @fig-type4variability d) may indicate that this chert is originally available as irregular (slightly oval) nodules within the parent rock.

Petrography results show this type is a massive wackestone composed mostly of micro-cryptocrystalline quartz (98%), with accessory macrocrystalline quartz and fibrous chalcedony. Allochems are iron oxides (c), opaques (vf), and bioclasts (c). Despite the macroscopically visible spicule-like white shapes, these fossils are not apparent in thin section. The porosity is low (1%).

```{r}
#| label: fig-type4variability
#| fig-cap: Chert macroscopic variability within type 4.

knitr::include_graphics("../figures/SOM/type4_macromeso.png")

@@

## Type 5

Type 5 corresponds to a massive and banded micro-cryptocrystalline quartz from a marine environment and possibly a shelf depositional environment (@fig-type5variability). The macroscopic and petrographic association of this type with the local cherts from the west Algarve basin, as well as the large size of the nodules and presence of parent rock, allow us to interpret this chert as local, possibly the Lower Jurassic although the exact outcrop is not identifiable.

This chert is fine to medium-grained, with a conchoidal fracture (albeit uneven at times). The quality is acceptable to low due to the lack of homogeneity and frequent fractures (@fig-type5variability c), as well as a coarser fabric in certain areas. It has a heterogeneous structure, frequently showing different characteristics within one single artifact. The colors are gray (5Y 6/1), light brownish gray (10YR 6/2), and pale brown (10YR 6/3). Fresh surfaces display slightly different colors: light gray (2.5Y 7/1) and gray (2.5Y 6/1). It has medium to dull luster, coarse to medium feel, and is opaque to sub-translucent; in the sub-translucent areas, light passes through to max. 4 mm of thickness. The type displays frequent patterns, mostly related to this heterogeneity, but also due to patination; these are splotched, banded, broad mottling, and speckling.

The cortex is from an outcrop, frequently rounded, but also irregular. Based on reference samples, the chert seems to be enclosed within the parent rock as irregular or tabular nodules (@fig-type5variability b). The cortex is thin (\~0.5 mm), with a sharp transition, and some artifacts show chunks of the parent rock attached to the cortex (@fig-type5variability a).

Based on the presence of bands, two sub-types were created: 5 (@fig-type5variability a-c) and 5B (@fig-type5variability d), the latter characterized by banding. Two thin sections were produced to characterize the macroscopic variability. The petrographic results show that type 5 is a wackestone, with a massive (type 5) or banded (type 5B) microstructure. They are mainly composed of micro-cryptocrystalline quartz (75% and 72%, respectively) and micrite (20% in both), and in smaller percentages macrocrystalline quartz and fibrous chalcedony. Allochems are unidentified opaques (vf), iron oxides (u), and bioclasts (c). The sub-types show larger differences in the porosity, with 5 showing 1% vug porosity while 5B shows 10% vug and moldic porosity, concentrated in the banded areas with more opaques.

Bioclasts are uncommon both macroscopically and in thin section: sponge spicules and round, unidentifiable fossils can be seen by the naked eye or under the microscope. In thin section however, several other unidentified fossils are present, some of which may be an ostracod (both in 5 and 5B) and a foraminifer (5).

```{r}
#| label: fig-type5variability
#| fig-cap: Chert macroscopic variability within type 5.

knitr::include_graphics("../figures/SOM/type5_macromeso.png")

@@

## Type 6

Type 6 corresponds to a massive cryptocrystalline quartz with mudstone texture (@fig-type6variability). The lack of local similar cherts, as well as identifiable fossils does not allow us to attribute a depositional environment or epoch/stage. However, rare spicules were identified; given the exclusivity of sponge's (Porifera) living environment to be marine [@flugelMicrofaciesCarbonateRocks2010], type 6 may be considered a chert formed in a marine environment. The similarities with cherts from Central Portugal indicates these cherts were formed in Cretaceous formations from the Lisbon and Estremadura area.

This chert is fine-grained and translucent (light passing through at 17 mm of thickness), with a conchoidal fracture and good quality. The structure is heterogeneous, with shiny to medium luster and semi-smooth to smooth feel. This type of translucent cherts shows a variety of colors: grayish brown (2.5Y 5/2), light brownish gray (10YR 6/2), brown (7.5 YR 4/2), reddish brown (2.5YR 4/3) and dark gray (10YR 4/1). The color distribution is variable; in some cases, the color seems to be related to the thickness of the sample; thicker parts show darker colors, while thinner parts show lighter colors. Alterations and inclusions also impact the color distribution and the discernment of natural versus the altered color. Alterations such as patina also seem to impact the translucency, which becomes opaque, and the luster, which becomes medium to dull.

@@

The cortex is rounded, from an outcrop source, and rarely rounded and smooth like a pebble. It is frequently around 0.5 to 1 mm thick and the transition is gradual.

```{r}
#| label: fig-type6variability
#| fig-cap: Chert macroscopic variability within type 6.

knitr::include_graphics("../figures/SOM/type6_macromeso.png")

@@

### Sub-type 6LLB

A sub-type named 6LLB (corresponding to the light brownish gray colors) was created during the macroscopic analysis process due to the presence of a type of artifacts with homogeneous characteristics (and refits) within the previous type (@fig-subtype6variability a-b).

This sub-type is characterized by white (10YR 8/1), light brownish gray (10YR 6/2), and brown (7.5YR 5/2) colors. It has a heterogeneous structure, with two fabrics: white and brown. The white fabric is located by the cortex and is opaque. It gradually turns to a light brownish gray which is sub-translucent (light passing through until \~10 mm), and to a brown which is translucent. The luster ranges from medium to dull, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is good. Patterns are laminated, splotched, and speckled. The patterns occur on the brown fabric, in a light brownish-gray color.

@@

```{r}
#| label: fig-subtype6variability
#| fig-cap: Chert macroscopic variability within the sub-type LLB (a-b) and 6E (c-d).

knitr::include_graphics("../figures/SOM/type6LLB_6E_variability.png")

@@

### Sub-type 6E

Similarly to the previous sub-type, a sub-type named 6E was created during the macroscopic analysis process due to the presence of a type of artifacts with homogeneous characteristics (and refits; @fig-subtype6variability c-d).

This sub-type is characterized by a heterogeneous structure. The colors are brown (10YR 5/3), gray (2.5Y 5/1), light gray (10YR 7/2) and white (10YR 8/1) The translucency is variable, some areas are translucent (light passing through at 8 mm of thickness), while the splotches are sub-translucent. The luster ranges from shiny to medium, the feel is smooth and the grain is fine. The fracture is conchoidal and the quality is acceptable, due to the presence of fractures. Patterns are splotches, broad mottling, spots, speckling, and laminated.

@@

Five thin sections were produced to better characterize the macroscopic variability of this type and sub-types. The results show no outstanding differences between the samples and sub-types. The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure. It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples, with one exception where it represents 91%), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz and micrite/sparite. Allochems are opaques (u), iron oxides (u) and bioclasts (r/u). Oxides and opaques are often concentrated in the edges of the sample, and may be related to post-depositional processes. Bioclasts are unidentifiable and poorly preserved, replaced by fibrous chalcedony and rarely by microcrystalline quartz. In a single sample (I20-3689) sponge spicules were identified. However, in sample H19-4074 (sub-type 11E), the absence of spicules or spicule-like shapes in the thin section suggests that they are an effect of white patina on the surface of the samples. Porosity is present between \<1-1%, of vug-type when identifiable.

## Type 7

Type 7 is a massive micro-cryptocrystalline, peloidal/oolitic packstone, from a marine environment and possibly a high-energy, shallow depositional environment (@fig-type7variability), due to the type, sorting, and preservation conditions of the ooids.
